Industrial, large-scale model predictive control with structured neural networks

### 2021-04-29 paper.tex
sync with published version.

%% The graphicx package provides the includegraphics command.
\usepackage{graphicx}
%% The amssymb package provides various useful mathematical symbols
\usepackage{amsmath}   % Extra math commands and environments from the AMS
\usepackage{amssymb}   % Special symbols from the AMS
\usepackage{amsthm}    % Enhanced theorem and proof environments from the AMS
\usepackage{latexsym}  % A few extra LaTeX symbols
\usepackage{array}
\usepackage{makecell}
\usepackage{units}
\usepackage{xcolor}
\usepackage[hidelinks=true]{hyperref}

@@

\usepackage{mpcsymbols}

@@

\begin{abstract}
The design of neural networks (NNs) is presented for treating large, linear
model predictive control (MPC) applications that are out of reach with available
quadratic programming (QP) solvers. First, we introduce a new feedforward
network architecture that enables practitioners to obtain offset-free
closed-loop performance with NNs. Second, we discuss the data generation
procedure to sample the state space relevant to training the NNs based on
anticipated online setpoint changes and plant disturbances. Third, we use the
input-to-state stability results available in the MPC literature and establish
robustness properties of NN controllers. Finally, we present illustrative
simulation studies on process control examples. We apply the NN design approach
and compare the performance with online QP based MPC on an industrial crude
distillation unit model with 252 states, 32 control inputs, and a control-sample
horizon length of 140. Parallel computing is used for data generation and
graphical processing units are used for network training. All the anticipated
plant operational scenarios with setpoints and disturbances that may change
during operation must be sampled for NN training. After the offline design
phase, NNs execute MPC three to five orders of magnitude faster than an
available QP solver with less than $1\%$ loss in the closed-loop performance.
\end{abstract}

@@

Model predictive control (MPC) is an online optimization based feedback control
technology. A dynamic model of the plant is used to make forecasts of the plant
measurements in response to the actuator movements, and an optimization problem
is solved in real time to determine the optimal actuator move to be applied to
the plant. For linear plant models, the optimization problem is a quadratic
program (QP). The development of efficient QP algorithms
\citep*{kouzoupis:frison:zanelli:diehl:2018, wright:2019} allowed practitioners
to apply MPC in the process industries \citep*{qin:badgwell:2003, lahiri:2017}.

An alternative strategy to real time optimization for the deployment of MPC is
to characterize the MPC feedback law as a piecewise affine function defined on
polyhedral partitions of the state space, and use the feedback law online
\citep*{bemporad:morari:dua:pistikopoulos:2002, seron:goodwin:dedona:2003}. The
real time computation in this approach is reduced to a table look-up in the
state space based on the current value of parameters in the QP such as the state
estimate and some steady-state targets that are computed for offset-free
control. The number of polyhedral partitions of the state space grows
exponentially with model dimensions, however, which prohibits the deployment of
this strategy on large-scale systems.

The MPC feedback law can be approximated using parametric functions such as
polynomials \citep*{kvasnica:lofberg:fikar:2011}, various types of piecewise
affine functions \citep*{bemporad:oliveri:poggi:storace:2011,
wen:ma:ydstie:2009}, and neural networks (NNs)
\citep*{cavagnari:magni:scattolini:1999}. Among the function approximators, NNs
that use the rectified linear unit (ReLU) as the activation function have gained
attention recently \citep*{chen:saulnier:atanasov:lee:kumar:pappas:morari:2018,
karg:lucia:2020, paulson:mesbah:2020, lovelett:dietrich:lee:kevrekidis:2020} to
approximate the MPC feedback law due to their ability to represent complex
piecewise affine functions \citep*{montufar:pascanu:cho:bengio:2014} and execute
MPC faster in real time than QP solvers. The proposed NN design approach in
these works is to generate training data for the NN by solving QPs offline for a
set of feasible states of the QP, use the collected data to train a standard
feedforward network, then use the trained NN as the feedback controller online.

To make progress in the size of control applications achievable with MPC, large
problems in which QP solvers fail to deliver the control in the available sample
time should be addressed. The scalability of the NN approach to such problems
with large model dimensions and control horizon length in the QP has not been
demonstrated to date. Two issues arise in these problems: i) In a large
dimensional state space, the entire set of feasible states of the QP cannot be
sampled densely for the NN training. This issue is a form of the well-known
curse of dimensionality. ii) One may sample the state space partially to avoid
this problem, but the time required for the data generation can still be
impractical due to the long time required to solve a QP for each sampled state.
We demonstrate in this article by case studies that this latter issue can be
resolved for a large size of problems with parallel computing, and the former
issue by sampling the state space for typical plant operational scenarios. The
following feature must be present in the large-scale application of interest for
the partial sampling scheme proposed in this article to be applicable: for a
particular operating mode of the plant, the number of frequently changing
setpoints and large magnitude disturbances must be reasonably small such that
the entire large dimensional state space is not visited during that operation
mode. The MPC feedback law can then be reliably approximated using NNs for the
different modes of plant operations driven by their respective set of setpoint
changes and disturbances.

From the above viewpoint, this article discusses the design of NNs as an
alternative to real time optimization in MPC, with an emphasis on large
applications that are challenging for QP solvers. The focus of this article is
on setpoint tracking MPC problems relevant for deployment in the process
industries. We present a new feedforward network architecture to obtain
offset-free closed-loop performance with the NNs. Additionally, we use the
input-to-state stability results available in the MPC literature, and establish
conditions under which NN controllers are robust to state estimation errors and
process disturbances. We present case studies on process control examples that
demonstrate the scalability of the proposed NN design approach.

The related work by \cite*{chen:wang:atanasov:kumar:morari:2019} demonstrates
the scalability of NNs to approximate the MPC feedback law on systems with state
dimensions up to 36. The setpoint tracking offset-free MPC problem and state
dimensions up to 250 considered in this article is not addressed in that work. 
\cite*{drgona:picard:kvasnica:helsen:2018}
take a different approach for the scalability of NN controllers, in which the
dimensionality of the parameters in the MPC QP is first reduced using principal
component analysis, and the NN controller is developed in a low dimensional
state space.

The rest of this article is organized as follows. In the next section, we
briefly review offset-free linear MPC. In Section \ref{sec:controller_design},
we discuss the NN controller design procedure. We present the robustness
property of NNs in Section \ref{sec:robustness}. Then, we present simulation
studies on two large process control examples in Section
\ref{sec:simulation_studies} and conclude at the end of this article.

@@

\subsection{Model}
We consider a linear discrete time model augmented with an integrating
disturbance model \citep{pannocchia:rawlings:2003}
\begin{align}\label{eq:ltimodel}
    x^+ = Ax + Bu + B_dd, \quad d^+ = d, \quad y = Cx + C_dd
\end{align}
in which $x \in \bbR^n$ is the state, $u \in \bbR^m$ is the control input, and
$d \in \bbR^d$ is the state of the disturbance model. The matrices $A \in
\bbR^{n \times n}$, $B \in \bbR^{n \times m}$,  
$C \in \bbR^{p \times n}$ denote the actuator to measurement model, and $B_d \in
\bbR^{n \times d}$, $C_d \in \bbR^{p \times d}$ is the disturbance model. The
objective of the disturbance model is to remove offset in the controlled plant
measurements at steady state operations and maintain the controlled variables at
their setpoints. We assume in the rest of this article that a Kalman filter can
be constructed to estimate the model state ($\xhat$) and disturbance ($\dhat$)
from measurements.

\subsection{Target selector}
Given the disturbance estimate, the input and controlled measurement setpoints,
we consider the following target selector QP
\begin{align}
    \min_{x_s, \ u_s}  \dfrac{1}{2} |u_{sp} - u_s|^2_{R_s}
\end{align}
subject to
\begin{align}  
    \begin{bmatrix}
        I -A & -B \\
        HC & 0
    \end{bmatrix} \begin{bmatrix}
        x_s \\
        u_s
    \end{bmatrix} &= \begin{bmatrix}
        B_d\dhat \\
        r_{sp} - HC_d\dhat
    \end{bmatrix} \\
    \underline{u} \leq u_s \leq \overline{u}
\end{align}
in which ($x_s, u_s$) is the target steady-state pair, $r=Hy$ are the controlled
measurements chosen as some subset or a linear combination of all the
measurements, ($\underline{u}, \overline{u}$) are the actuator bounds, and
($u_{sp}, r_{sp}$) are the input and controlled measurement setpoints. The
equality constraint $r_{sp} = H(Cx_s + C_d\dhat)$ can be difficult to satisfy
exactly in real time depending on the controlled measurement setpoint and
disturbance estimate. In this case, the hard setpoint equality constraint can be
relaxed and moved to the stage cost such that the target selector computes a
steady state to minimize the offset in the controlled measurements. We assume
for the simulation studies in this article that the input setpoint ($u_{sp}$) is
fixed at some chosen steady state and only the controlled measurement setpoint
($r_{sp}$) changes in real time.

\subsection{Regulator}
Based on the state estimate and target steady-state pair, the following QP is
solved to determine the actuator move to be applied to the plant
\begin{align} \label{eqn:regulator}
    V_N(\tilde{x}(0), \mathbf{\tilde{u}}) = \dfrac{1}{2}\sum_{k=0}^{N-1} & 
    \Big(|\tilde{x}(k)|^2_Q  + |\tilde{u}(k)|^2_R \Big) + 
    \dfrac{1}{2}|\tilde{x}(N)|^2_P
\end{align}
subject to 
\begin{align}
    \tilde{x}^+ &= A\tilde{x} + B\tilde{u}, \quad \tilde{x}(0) = \xhat - x_s \\
    &\underline{u} \leq \tilde{u} + u_s \leq \overline{u}
    \label{eqn:regulator_constraints}
\end{align}
in which $\xhat$ is the state estimate after the current measurement, $N$ is the
control horizon length, and ($\tilde{x}, \tilde{u}$) are the state and control
in deviation from the current steady-state targets ($x_s, u_s$). The penalty
matrix $P$ is chosen as the optimal cost-to-go matrix of the unconstrained
infinite horizon linear quadratic problem. The decision variable in this QP is
the control sequence $\mathbf{\tilde{u}} = [\tilde{u}(0)', \tilde{u}(1)', ...,
\tilde{u}(N-1)']'$. The first move of the optimal solution is denoted as the MPC
feedback law, $\kappa_N(\xhat, x_s, u_s) = \mathbf{\tilde{u}}^0(0;\xhat, x_s,
u_s) + u_s$, which is applied to the plant. This feedback law is a function of
the state estimate and target steady-state pair. A rate-of-change penalty on the
control can be imposed in the QP objective by augmenting the dynamic model with
a surrogate state for the control as discussed in \cite*{rao:rawlings:1999}. In
this case, the control injected at the previous timestep ($u_{-1}$) is supplied
to the QP, and the feedback law becomes a function of the previous control as
well in addition to the state estimate and target steady-state pair.

Several algorithms have been proposed in the literature over the years to solve
the above MPC regulator QP. \cite*{kouzoupis:frison:zanelli:diehl:2018} and
\cite*{wright:2019} give recent reviews on the development of convex
optimization algorithms to solve this linear MPC QP. For the simulation studies
in this article, we eliminate the state trajectory from the set of decision
variables and formulate a dense QP. The decision variable in the dense QP is the
future control trajectory $\mathbf{\tilde{u}}$, and the QP solver CVXOPT
\citep*{vandenberghe:2010}, which is tailored for dense problems is used for NN
data generations and timing comparisons with online QP based MPC. Any
improvement in the QP formulation and solution algorithm is advantageous for
both the online QP based MPC and NN approaches as fast QP solvers reduce the
offline data generation time required for the design of NNs.

@@

\subsection{Structured neural network}
An intuitive approach is to build a feedforward NN that takes the triple ($x,
x_s, u_s$) as its input and outputs a control that is close to the optimal
control. This strategy has been proposed in \cite*{karg:lucia:2020} and 
\cite*{chen:saulnier:atanasov:lee:kumar:pappas:morari:2018}
for a single fixed steady state. The approach has not been demonstrated to scale
to the size of model dimensions and the setpoint tracking offset-free MPC
problem considered in this article. An issue with this strategy is that the NN
has no knowledge about the MPC feedback law, and can require large amounts of
training data in large state dimensions to obtain a reasonable controller. At a
minimum, the MPC feedback law is such that at the origin, the control action
taken by the MPC controller is zero. For the setpoint tracking MPC problem
considered in this article, the NN control law must be such that at steady-state
operations, the NN maintains the plant at the desired steady state such that the
controlled measurements are at their setpoints. This structure implies $u_s =
\kappa_{NN}(x = x_s, x_s, u_s)$, in which we use $\kappa_{NN}(\cdot)$ to denote
the NN control law. To incorporate this information in a NN, we introduce the
following architecture
\begin{align}
z_0 &= \begin{bmatrix}
    x', & x_s', & u_s', & x_s', & x_s', & u_s' \end{bmatrix}'\notag \\
    f_i(z_{i-1}) &= \begin{bmatrix}
        W_i & 0 \notag \\
        0 & W_i
          \end{bmatrix}z_{i-1}  + \begin{bmatrix}
            b_i \\
            b_i
            \end{bmatrix} \notag\\
    z_i &= g(f_i(z_{i-1})), \textnormal{for} \ i \in \bbI_{1:h} \notag \\
    u &= u_s + \begin{bmatrix}
  W_{h + 1}, & -W_{h + 1} \\
    \end{bmatrix} z_h
\label{eq:NNstruct}
\end{align}
in which $g(a) = \textnormal{max} (0, a)$, is the ReLU operation applied
element-wise on its vector input, $z_i$ is the output of each hidden layer of
the network, $z_0$ is the input to the network, $h$ is the number of hidden
layers, and $i \in \bbI_{1:h}$. The parameters to be optimized using training
data are ($W_i, b_i$) and $W_{h+1}$, which are matrices of appropriate
dimensions. At steady state, when $x=x_s$ in $z_0$, \eqref{eq:NNstruct} outputs
$u_s$ regardless of the choice of weights and the activation function, as the
output of all hidden layers ($z_i$) have same subvector repeated in the upper
and lower half and the term $\begin{bmatrix} W_{h + 1}, & -W_{h+ 1} \\
      \end{bmatrix} z_h$ equals zero. The NN structure \eqref{eq:NNstruct} is
equivalent to the difference in outputs of two standard NNs with inputs ($x,
x_s, u_s$) and ($x_s, x_s, u_s$) respectively, with $u_s$ added to the final
output. The software tensorflow \citep{abadi:agarwal:barham:brevdo:et-al:2015}
is used for NN training purposes in this article, and the structured
architecture can be built with symbolic differentiation tools available in
tensorflow. When a rate-of-change penalty on the control is included in the
regulator QP, the input to the structured NN is modified to 
\begin{align*}
    z_0 &= 
    \begin{bmatrix}
    x', & u_{-1}', & x_s', & u_s', & x_s', & u_s', & x_s', & u_s' \\
    \end{bmatrix}'
\end{align*}
in which $u_{-1}$ is the previous control applied to the plant and the
dimensions of the matrices ($W_i, b_i$) and $W_{h+1}$ are appropriately
adjusted. Even after training, the control produced by the NN may not satisfy
the hard actuator constraints, and the saturation function
($\textnormal{sat}(\cdot, \underline{u}, \overline{u})$) is applied to the
output of the NN.

@@

For one fixed steady state, the set of feasible states for the MPC QP defined by
\eqref{eqn:regulator}-\eqref{eqn:regulator_constraints} is the entire
$n$-dimensional state space $\bbR^n$. Additionally, the target steady state may
change during the plant operation depending on the setpoint and disturbance
estimate. The domain of the MPC control law ($\kappa_N(x, x_s, u_s)$) is
therefore the state space $\bbR^{n}$ and the possible set of steady-state target
pairs. Sampling this entire domain of the MPC control law for NN training is
difficult in large dimensions, and we show in Section
\ref{sec:simulation_studies} that it is also not required for practical
applications.

From a plant operations perspective, a majority of the large-scale petrochemical
plants operate in a relatively small number of operating regimes or scenarios.
Each operational scenario is driven by a selected few controlled measurement
setpoints that depend on product demands and some large magnitude disturbances
may change frequently, while the setpoints for the other measurements remain
constant for long periods of time. The states and steady-state targets spanned
in a closed-loop operation per scenario is thereby considerably less than the
entire domain of the MPC control law. This typical operation of large chemical
plants can be exploited, and we propose the following data generation procedure
to perform offline simulations using the model to collect the operationally
relevant set of ($x, x_s, u_s, \kappa_N(\cdot)$) for NN training.

@@

During the plant operation, it is expected that the NN does not encounter the
same states and steady-state target pairs used for the training. But similar
values are expected since an adequate model and appropriate range of anticipated
setpoints and plant disturbances are used for the data generation. The NN during
the closed-loop operation produces an interpolated actuator move that is applied
to the plant.

As we consider large MPC problem sizes in this article for the NN controller,
the data generation step can take impractical amounts of time if performed
serially due to the long time required to solve each QP (minutes). For these
large problems, several PRBS signals of the setpoints and disturbances are
generated and multiple simulations are performed in parallel over several cores
in a CPU and across multiple CPUs as well. The data generated from all the
simulations are collected and used for the NN training.

@@

We now present the robustness properties of NN controllers. The optimal MPC
controller designed for a nominal linear system ignoring disturbances is known
to be inherently robust to bounded state estimation errors and disturbances
\citep*{heath:wills:2005, pannocchia:rawlings:wright:2011}. The approximation
error of the MPC feedback law by a NN can be viewed as an additional
disturbance, and the existing input-to-state stability results
\citep*{sontag:wang:1995} are used to establish the allowable approximation
error in the NN for the NN controller to be robust to disturbances. Theorem
\ref{thm:nnrobustness} states this allowable approximation error and specifies
the size of disturbances for which the NN controller is robust. This result also
holds for other approximate MPC controllers, which use different parametric
functions to approximate the optimal control law such as polynomials and other
piecewise affine functions.

>>>>

%% The graphicx package provides the includegraphics command.
\usepackage{graphicx}
%% The amssymb package provides various useful mathematical symbols
\usepackage{amsmath}   % Extra math commands and environments from the AMS
\usepackage{amssymb}   % Special symbols from the AMS
\usepackage{amsthm}    % Enhanced theorem and proof environments from the AMS
\usepackage{latexsym}  % A few extra LaTeX symbols
\usepackage{array}
\usepackage{makecell}
\usepackage{units}
\usepackage{xcolor}
\usepackage[hidelinks=true]{hyperref}
\usepackage{natbib}

%\biboptions{authoryear,round}

@@

% Define shortcuts in MPCSymbols.
\newcommand{\bbI}{\mathbb{I}}
\newcommand{\bbR}{\mathbb{R}}
\newcommand{\norm}[1]{\left| #1 \right|}
\newcommand{\xhat}{\hat{x}}
\newcommand{\dhat}{\hat{d}}
\newcommand{\xseq}{\mathbf{x}}
\newcommand{\bbX}{\mathbb{X}}
\newcommand{\bbW}{\mathbb{W}}

@@

\begin{abstract}
The design of neural networks (NNs) is presented for treating large, linear
model predictive control (MPC) applications that are out of reach with available
quadratic programming (QP) solvers. First, we introduce a new feedforward
network architecture that enables practitioners to obtain offset-free
closed-loop performance with NNs. Second, we discuss the data generation
procedure to sample the state space relevant to training the NNs based on
anticipated online setpoint changes and plant disturbances. Third, we use the
input-to-state stability results available in the MPC literature and establish
robustness properties of NN controllers. Finally, we present illustrative
simulation studies on process control examples. We apply the NN design approach
and compare the performance with online QP based MPC on an industrial crude
distillation unit model with 252 states, 32 control inputs, and a control-sample
horizon length of 140. Parallel computing is used for data generation and
graphical processing units are used for network training. Anticipated plant
operational scenarios with setpoints and disturbances that may change during
operation must be sampled for NN training. After the offline design phase, NNs
execute MPC three to five orders of magnitude faster than an available QP solver
with less than $1\%$ loss in the closed-loop performance.
\end{abstract}

@@

Model predictive control (MPC) is an online optimization based feedback control
technology. A dynamic model of the plant is used to make forecasts of the plant
measurements in response to the actuator movements, and an optimization problem
is solved in real time to determine the optimal actuator move to be applied to
the plant. For linear plant models, the optimization problem is a quadratic
program (QP). The development of efficient QP algorithms
\citep{kouzoupis:frison:zanelli:diehl:2018, wright:2019} allowed practitioners
to apply MPC in the process industries \citep*{qin:badgwell:2003, lahiri:2017}.

An alternative strategy to real time optimization for the deployment of MPC is
to characterize the MPC feedback law as a piecewise affine function defined on
polyhedral partitions of the state space, and use the feedback law online
\citep{bemporad:morari:dua:pistikopoulos:2002, seron:goodwin:dedona:2003}. The
real time computation in this approach is reduced to a table look-up in the
state space based on the current value of parameters in the QP such as the state
estimate and some steady-state targets that are computed for offset-free
control. The number of polyhedral partitions of the state space grows
exponentially with model dimensions, however, which prohibits the deployment of
this strategy on large-scale systems.

The MPC feedback law can be approximated using parametric functions such as
polynomials \citep{kvasnica:lofberg:fikar:2011}, various types of piecewise
affine functions \citep{bemporad:oliveri:poggi:storace:2011,
wen:ma:ydstie:2009}, and neural networks (NNs)
\citep{cavagnari:magni:scattolini:1999}. Among the function approximators, NNs
that use the rectified linear unit (ReLU) as the activation function have gained
attention recently \citep{chen:saulnier:atanasov:lee:kumar:pappas:morari:2018,
karg:lucia:2020, paulson:mesbah:2020, lovelett:dietrich:lee:kevrekidis:2020} to
approximate the MPC feedback law due to their ability to represent complex
piecewise affine functions \citep{montufar:pascanu:cho:bengio:2014} and execute
MPC faster in real time than QP solvers. The proposed NN design approach in
these works is to generate training data for the NN by solving QPs offline for a
set of feasible states of the QP, use the collected data to train a standard
feedforward network, then use the trained NN as the feedback controller online.

To make progress in the size of control applications achievable with MPC, large
problems in which QP solvers fail to deliver the control in the available sample
time should be addressed. The scalability of the NN approach to such problems
with large model dimensions and control horizon length in the QP has not been
demonstrated to date. Two issues arise in these problems: i) In a large
dimensional state space, the entire set of feasible states of the QP cannot be
sampled densely for the NN training. This issue is a form of the well-known
curse of dimensionality. ii) One may sample the state space partially to avoid
this problem, but the time required for the data generation can still be
impractical due to the long time required to solve a QP for each sampled state.
We demonstrate in this article by case studies that the issue (ii) can be
resolved for many size of problems with parallel computing, and the issue (i) by
sampling the state space for typical plant operational scenarios. The following
feature must be present in the large-scale application of interest for the
partial sampling scheme proposed in this article to be applicable: For a
particular operating mode of the plant, the number of frequently changing
setpoints and large magnitude disturbances must be small enough such that only a
small fraction of the entire large dimensional state space is visited during
that operation mode. The MPC feedback law can then be reliably approximated
using NNs for the different modes of plant operations, each driven by their
respective set of setpoint changes and disturbances.

From the above viewpoint, this article discusses the design of NNs as an
alternative to real time optimization in MPC, with an emphasis on large
applications that are challenging for QP solvers. We focus on setpoint tracking
MPC problems relevant to the process industries. We present a new feedforward
network architecture to obtain offset-free closed-loop performance with the NNs.
Additionally, we use the input-to-state stability results available in the MPC
literature, and establish conditions under which NN controllers are robust to
state estimation errors and process disturbances. We present case studies on
process control examples that demonstrate the scalability of the proposed NN
design approach.

The related work of \cite{chen:wang:atanasov:kumar:morari:2019} demonstrates the
scalability of NNs to approximate the MPC feedback law on systems with state
dimensions up to 36, on the control problem of regulation to the origin. By
contrast, our paper considers setpoint tracking offset-free MPC problem and
state dimensions up to 250. \cite{drgona:picard:kvasnica:helsen:2018} take a
different approach for the scalability of NN controllers, in which the
dimensionality of the parameters in the MPC QP is first reduced using principal
component analysis, and the NN controller is developed in a low dimensional
state space.

The rest of this article is organized as follows. In the next section, we
briefly review offset-free linear MPC. In Section \ref{sec:controller_design},
we discuss the NN controller design procedure. We present the robustness
property of NNs in Section \ref{sec:robustness}. Section
\ref{sec:simulation_studies} presents simulation studies on two large process
control examples and some concluding thoughts appear in Section
\ref{sec:conclusion}.

@@

\subsection{Model}
We consider a linear discrete time model augmented with an integrating
disturbance model \citep{pannocchia:rawlings:2003}:
\begin{align}\label{eq:ltimodel}
    x^+ = Ax + Bu + B_dd, \quad d^+ = d, \quad y = Cx + C_dd,
\end{align}
in which $x \in \bbR^n$ is the state, $u \in \bbR^m$ is the control input, and
$d \in \bbR^d$ is the state of the disturbance model. The matrices $A \in
\bbR^{n \times n}$, $B \in \bbR^{n \times m}$,  
$C \in \bbR^{p \times n}$ denote the actuator to measurement model; and $B_d \in
\bbR^{n \times d}$, $C_d \in \bbR^{p \times d}$ is the disturbance model. The
objective of the disturbance model is to remove offset in the controlled plant
measurements at steady state operations and maintain the controlled variables at
their setpoints. We assume in the rest of this article that a Kalman filter can
be constructed to estimate the model state ($\xhat$) and disturbance ($\dhat$)
from measurements.

\subsection{Target selector}
Given the disturbance estimate, the input and controlled measurement setpoints,
we consider the following target selector QP:
\begin{align}
    \min_{x_s, \ u_s}  \dfrac{1}{2} |u_{sp} - u_s|^2_{R_s}
\end{align}
subject to
\begin{align}  
    \begin{bmatrix}
        I -A & -B \\
        HC & 0
    \end{bmatrix} \begin{bmatrix}
        x_s \\
        u_s
    \end{bmatrix} &= \begin{bmatrix}
        B_d\dhat \\
        r_{sp} - HC_d\dhat
    \end{bmatrix}, \\
    \underline{u} \leq u_s \leq \overline{u},
\end{align}
in which ($x_s, u_s$) is the target steady-state pair, $r=Hy$ are the controlled
measurements chosen as some subset or a linear combination of all the
measurements, ($\underline{u}, \overline{u}$) are the actuator bounds, and
($u_{sp}, r_{sp}$) are the input and controlled measurement setpoints. The
equality constraint $r_{sp} = H(Cx_s + C_d\dhat)$ can be difficult to satisfy
exactly in real time depending on the controlled measurement setpoint and
disturbance estimate. In this case, the hard setpoint equality constraint can be
relaxed and moved to the stage cost, such that the target selector computes a
steady state to minimize the offset in the controlled measurements. We assume
for the simulation studies in this article that the input setpoint ($u_{sp}$) is
fixed at some chosen steady state and only the controlled measurement setpoint
($r_{sp}$) changes in real time.

\subsection{Regulator}
Based on the state estimate and target steady-state pair, the following QP is
solved to determine the actuator move to be applied to the plant:
\begin{align} \label{eqn:regulator}
    V_N(\tilde{x}(0), \mathbf{\tilde{u}}) = \dfrac{1}{2}\sum_{k=0}^{N-1} & 
    \Big(|\tilde{x}(k)|^2_Q  + |\tilde{u}(k)|^2_R \Big) + 
    \dfrac{1}{2}|\tilde{x}(N)|^2_P
\end{align}
subject to 
\begin{align}
    \tilde{x}^+ &= A\tilde{x} + B\tilde{u}, \quad \tilde{x}(0) = \xhat - x_s, \label{eqn:model_eq} \\
    &\underline{u} \leq \tilde{u} + u_s \leq \overline{u},
    \label{eqn:regulator_constraints}
\end{align}
in which $\xhat$ is the state estimate after the current measurement, $N$ is the
control horizon length, and ($\tilde{x}, \tilde{u}$) are the state and control
in deviation from the current steady-state targets ($x_s, u_s$). The penalty
matrix $P$ is chosen as the optimal cost-to-go matrix of the unconstrained
infinite horizon linear quadratic problem. The decision variable in this QP is
the control sequence $\mathbf{\tilde{u}} = [\tilde{u}(0)', \tilde{u}(1)', ...,
\tilde{u}(N-1)']'$, and the state sequence $\tilde{\xseq}$ is fully determined
by the model equality constraints \eqref{eqn:model_eq}. The first move of the
optimal solution is denoted as the MPC feedback law, $\kappa_N(\xhat, x_s, u_s)
= \mathbf{\tilde{u}}^0(0;\xhat, x_s, u_s) + u_s$, which is applied to the plant.
This feedback law is a function of the state estimate and target steady-state
pair. A rate-of-change penalty on the control can be imposed in the QP objective
by augmenting the dynamic model with a surrogate state for the control as
discussed in \cite{rao:rawlings:1999}. In this case, the control injected at the
previous timestep ($u_{-1}$) is supplied to the QP, and the feedback law becomes
a function of the previous control, in addition to the state estimate and target
steady-state pair.

Several algorithms have been proposed in the literature over the years to solve
the above MPC regulator QP. \cite{kouzoupis:frison:zanelli:diehl:2018} and
\cite{wright:2019} give recent reviews on the development of convex optimization
algorithms to solve this linear MPC QP. For the simulation studies in this
article, we eliminate the state trajectory from the set of decision variables
and formulate a dense QP. The decision variable in the dense QP is the future
control trajectory $\mathbf{\tilde{u}}$, and the QP solver CVXOPT
\citep{vandenberghe:2010}, which is tailored for dense problems is used to
generate data for NN training and for timing comparisons with online QP based
MPC. Any improvement in the QP formulation and solution algorithm is
advantageous for both the online QP based MPC and NN approaches, as fast QP
solvers reduce the offline data generation time required for the design of NNs.

@@

\subsection{Structured neural network}
An intuitive approach is to build a feedforward NN that takes the triple ($x,
x_s, u_s$) as its input and outputs a control that is close to the optimal
control. This strategy has been proposed in \cite{karg:lucia:2020} and 
\cite{chen:saulnier:atanasov:lee:kumar:pappas:morari:2018}
for a single fixed steady state. The approach has not been demonstrated to scale
to the size of model dimensions and the setpoint tracking offset-free MPC
problem considered in this article. An issue with this strategy is that the NN
has no knowledge about the MPC feedback law, and can require large amounts of
training data in large state dimensions to obtain a reasonable controller. At a
minimum, the MPC feedback law is such that at the origin, the control action
taken by the MPC controller is zero. For the setpoint tracking MPC problem
considered in this article, the NN control law must be such that at steady-state
operations, the NN maintains the plant at the desired steady state such that the
controlled measurements are at their setpoints. This structure implies $u_s =
\kappa_{NN}(x = x_s, x_s, u_s)$, in which we use $\kappa_{NN}(\cdot)$ to denote
the NN control law. To incorporate this information in a NN, we introduce the
following architecture:
\begin{align}
z_0 &= \begin{bmatrix}
    x', & x_s', & u_s', & x_s', & x_s', & u_s' \end{bmatrix}', \notag \\
    f_i(z_{i-1}) &= \begin{bmatrix}
        W_i & 0 \notag \\
        0 & W_i
          \end{bmatrix}z_{i-1}  + \begin{bmatrix}
            b_i \\
            b_i
            \end{bmatrix}, \notag\\
    z_i &= g(f_i(z_{i-1})), \textnormal{for} \ i \in \bbI_{1:h}, \notag \\
    u &= u_s + \begin{bmatrix}
  W_{h + 1}, & -W_{h + 1} \\
    \end{bmatrix} z_h,
\label{eq:NNstruct}
\end{align}
in which $g(a) = \textnormal{max} (0, a)$, is the ReLU operation applied
element-wise on its vector input, $z_i$ is the output of each hidden layer of
the network, $z_0$ is the input to the network, $h$ is the number of hidden
layers, and $i \in \bbI_{1:h}$. The parameters to be optimized using training
data are ($W_i, b_i$) and $W_{h+1}$, which are matrices of appropriate
dimensions. At steady state, when $x=x_s$ in $z_0$, \eqref{eq:NNstruct} outputs
$u_s$ regardless of the choice of weights and the activation function, as the
output of all hidden layers ($z_i$) have same subvector repeated in the upper
and lower half and the term $\begin{bmatrix} W_{h + 1}, & -W_{h+ 1} \\
      \end{bmatrix} z_h$ equals zero. The NN structure \eqref{eq:NNstruct} is
equivalent to the difference in the outputs of a standard feedforward NN with
inputs ($x, x_s, u_s$) and ($x_s, x_s, u_s$) respectively, with $u_s$ added to
the final output. The software tensorflow
\citep{abadi:agarwal:barham:brevdo:et-al:2015} is used for NN training; the
structured architecture can be built with symbolic differentiation tools
available in tensorflow.

When a rate-of-change penalty on the control is included in the regulator QP,
the input to the structured NN is modified to:
\begin{align*}
    z_0 &= 
    \begin{bmatrix}
    x', & u_{-1}', & x_s', & u_s', & x_s', & u_s', & x_s', & u_s' \\
    \end{bmatrix}',
\end{align*}
in which $u_{-1}$ is the previous control applied to the plant and the
dimensions of the matrices ($W_i, b_i$) and $W_{h+1}$ are appropriately
adjusted. Even after training, the control produced by the NN may not satisfy
the hard actuator constraints, so the saturation function
($\textnormal{sat}(\cdot, \underline{u}, \overline{u})$) is applied to the
output of the NN.

@@

For one fixed steady state, the set of feasible states for the MPC QP defined by
\eqref{eqn:regulator}-\eqref{eqn:regulator_constraints} is the entire
$n$-dimensional state space $\bbR^n$. Additionally, the target steady state may
change during plant operation, depending on the setpoint and disturbance
estimate. The domain of the MPC control law ($\kappa_N(x, x_s, u_s)$) is
therefore the state space $\bbR^{n}$ and the possible set of steady-state target
pairs. Sampling this entire domain of the MPC control law for NN training is
difficult in large dimensions; we show in Section \ref{sec:simulation_studies}
that it is not required for practical applications.

A majority of large-scale petrochemical plants operate in a relatively small
number of operating regimes or scenarios. Each operational scenario is driven by
a selected few controlled measurement setpoints that depend on product demands
and some large magnitude disturbances that may change frequently, while the
setpoints for other measurements remain constant for long periods of time. The
states and steady-state targets spanned in a closed-loop operation per scenario
is thereby considerably less than the entire domain of the MPC control law.
These characteristics of typical operation of large chemical plants can be
exploited, and we propose the following data generation procedure to perform
offline simulations using the model to collect the operationally relevant set of
($x, x_s, u_s, \kappa_N(\cdot)$) for NN training.

@@

During plant operation, it is expected that the NN will encounter states and
steady-state target pairs not used in training. But similar values are expected,
since an adequate model and appropriate range of anticipated setpoints and plant
disturbances are used for the data generation. During the closed-loop operation,
the NN produces an interpolated actuator move that is applied to the plant.

As we consider large MPC problem sizes in this article for the NN controller,
the data generation step can take impractical amounts of time if performed
serially, due to the long time (minutes) required to solve each QP. For these
large problems, several PRBS signals of the setpoints and disturbances are
generated and multiple simulations are performed in parallel over several cores
in a CPU and across multiple CPUs as well. The data generated from all the
simulations are collected and used for the NN training.

@@

We now present the robustness properties of NN controllers. The optimal MPC
controller designed for a nominal linear system ignoring disturbances is known
to be inherently robust to bounded state estimation errors and disturbances
\citep{heath:wills:2005, pannocchia:rawlings:wright:2011}. The approximation
error of the MPC feedback law by a NN can be viewed as an additional
disturbance, and existing input-to-state stability results
\citep{sontag:wang:1995} can be used to establish the allowable approximation
error in the NN for the NN controller to be robust to disturbances. Theorem
\ref{thm:nnrobustness} below states this allowable approximation error and
specifies the size of disturbances for which the NN controller is robust. This
result also holds for other approximate MPC controllers, which use different
parametric functions to approximate the optimal control law, such as polynomials
and other piecewise affine functions.
